---
title: "Assignment 1"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#ASSIGNMENT
Packages

```{r}
library(phytools)
library(geiger)
```

##Question 1
Load data and use both tests to check the tree is fully bifurcated

```{r}
snake.tree<-read.tree("homalops.phy") ; snake.tree
is.binary(snake.tree)
```

Run ltt

```{r}
snake.obj<-ltt(snake.tree,plot = FALSE) ; snake.obj
snake.gamma <- snake.obj$gamma ; snake.gamma
```

GAMMA: -3.2411  
p-val: 0.0012

##Question 2
The gamma value is conclusively negative which means there are more nodes/divergences earlier in the tree. This would be reason for further study to determine if an adaptive radiation event has occurred to explain the rise of more lineages farther back in time than in the more recent past.

##Question 3
Determine if the observed gamma is due to incomplete sampling.

Create a set of simulated lineages and see how tree gamma varies when 15 lineages are excluded each time and plot a histogram of the resulting values. 

```{r}
age <- 22
richness <- 34
missing <- 13

snakebirth =  (log(richness) - log(2))/age ; snakebirth
num_simulations<-200 

# For loop to simulate populations 
# COPIED FROM MATERIAL PROVIDED BY PROF. MIKE ALFARO
g1_null<-numeric(num_simulations) #g1_null will hold the simulated gamma values
for(i in 1:num_simulations) {
  sim.bdtree(snakebirth, d=0, stop = "taxa", n=richness)->sim_tree 
  drop.random(sim_tree, missing)->prune # prune down to the # of taxa in the phylogeny
  gammaStat(prune)->g1_null[i]
}

hist(g1_null)
arrows(snake.gamma, 40, snake.gamma, 0, col="red", lwd=2) 
```

The arrow indicating the observed gamma value seems to be well into the tails and therefore possibly significant. We will now determine the p value.

```{r}
smallerNull<-g1_null<=snake.gamma
count<-sum(smallerNull)
mccr_pval<-(count+1)/(num_simulations+1) ; mccr_pval
```

p-val: 0.00995  

On the basis of the MCCR test, I conclude the rate of speciation slowing as time increases.

##Question 4
Fit birth-death model and report rates:

```{r}
fitbd <- birthdeath(snake.tree)
bd(fitbd)
```

Birth Rate: 0.0684 
Death Rate: 0.0000 (The species are all still extant so none have gone extinct)

##Question 5
####(1) Clade description and source  
Number of Tips: 50  
Number of Species in Clades: 54  
Mustelidae data set downloaded from:  
http://10ktrees.nunn-lab.org/Carnivora/downloadTrees.php

Load tree and check it is fully bifurcated.

```{r}
prac.tree<-read.nexus("consensusTree_10kTrees_Carnivora_Version1.nex") ; prac.tree
is.binary(prac.tree)
```

####(2) Fit birth death model and return b and d

```{r}
fitbd <- birthdeath(prac.tree)
bd(fitbd)
```

Birth Rate: 0.0769  
Death Rate: 0.0000  
Again, all species are extant

####(3) Perform MCCR test and analyse  
Run ltt and get observed gamma

```{r}
prac.obj<-ltt(prac.tree,plot = FALSE) ; prac.obj
prac.gamma <- prac.obj$gamma ; prac.gamma
```

Simulate incomplete sampling

```{r}
age <- 15
richness <- 54
pracbirth =  (log(richness) - log(2))/age ; pracbirth
missing <- 4
num_simulations<-200 
# For loop to simulate populations 
# COPIED FROM MATERIAL PROVIDED BY PROF. MIKE ALFARO
g1_null<-numeric(num_simulations) #g1_null will hold the simulated gamma values
for(i in 1:num_simulations) {
  sim.bdtree(pracbirth, d=0, stop = "taxa", n=richness)->sim_tree 
  drop.random(sim_tree, missing)->prune # prune down to the # of taxa in the phylogeny
  gammaStat(prune)->g1_null[i]
}


hist(g1_null)
arrows(prac.gamma, 40, prac.gamma, 0, col="red", lwd=2) 
```

Seems to be somewhat into the tails. Looks possibily significant.  

Generate p-value of observed gamma

```{r}
smallerNull<-g1_null<=prac.gamma
count<-sum(smallerNull)
mccr_pval<-(count+1)/(num_simulations+1) ; mccr_pval
```

p-val: 0.0448    

Interpratation:   
The gamma value does seem to be significant compared to simulations suggesting a slight skew of the nodes towards the beginning of the tree which also suggests the rate of speciation declined over time.
