---
title: "Lab 1"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#TUTORIAL
##Exercise 1
####PACKAGES
```{r}
library(phytools)
```

####LOAD DATA
```{r}
darter.tree<-read.tree("etheostoma_percina_chrono.tre")
```

####PLOT TREE
```{r}
plotTree(darter.tree,ftype="i",fsize=0.4,type="fan",lwd=1)
```

####LTT PLOT 
```{r}
obj<-ltt(darter.tree,log.lineages=FALSE) ; obj
```

####EXAMINE DATA SET 
```{r}
darter.tree
```
A bifurcated tree would have 200 internal nodes (n_tips-1).
Another check of whether the tree is fully bifurcating...
```{r}
is.binary(darter.tree)
```

####FULLY BIFURCATE TREE 
```{r}
darter.tree<-multi2di(darter.tree) ; darter.tree
is.binary(darter.tree)
```
Have now forced multifurcating nodes to bifurcate

####STORE LTT
```{r}
obj<-ltt(darter.tree,plot = FALSE) ; obj
```

####PLOT FULLY BIFURCATED DATA SET AND OVERLAY TREE ONTO LTT PLOT
```{r}
plot(obj,log.lineages=FALSE,main="LTT plot for darters")
plot(obj,show.tree=TRUE,log.lineages=FALSE,main="LTT plot for darters")
```

####PLOT LOG OF LINEAGE NUMBER AGAINST TIME AND ADD LINE SHOWING PREDICTION UNDER PURE BIRTH MODEL
```{r}
plot(obj,log.lineages=FALSE,log="y",main="LTT plot for darters",
     ylim=c(2,Ntip(darter.tree)))
h<-max(nodeHeights(darter.tree))
x<-seq(0,h,by=h/100)
b<-(log(Ntip(darter.tree))-log(2))/h
lines(x,2*exp(b*x),col="red",lty="dashed",lwd=2)
```

####SIMULATED LINEAGE NUMBERS THROUGH TIME WITH ORIGINAL DATA
```{r}
trees<-pbtree(b=b,n=Ntip(darter.tree),t=h,nsim=100,method="direct",
              quiet=TRUE)
obj<-ltt(trees,plot=FALSE)
plot(obj,col="grey",main="LTT of darters compared to simulated LTTs")
lines(c(0,h),log(c(2,Ntip(darter.tree))),lty="dashed",lwd=2,col="red")
ltt(darter.tree,add=TRUE,lwd=2)
```

####95% CONFIDENCE INTERVAL FOR LTT FROM A SET OF TREES 
```{r}
ltt95(trees,log=TRUE)
title(main="LTT of darters compared to simulated LTTs")
ltt(darter.tree,add=TRUE,log.lineages=FALSE,col="red",lwd=2)
```

##Exercise 2
####FIT BIRTH-DEATH MODEL AND RETURN RATES
```{r}
fitbd <- birthdeath(darter.tree)
bd(fitbd)
```

##Exercise 3
####PLOT HISTOGRAM OF SIMULATED GAMMA VALUES AND RETURN SUMMARY STATS
```{r}
g<-sapply(trees,function(x) ltt(x,plot=FALSE)$gamma)
hist(g,main=expression(paste("Distribution of ",gamma," from simulation")))
mean(g)
var(g)
```

####TESTING HYPOTHESES
```{r}
obj<-ltt(darter.tree,plot=FALSE)
print(obj)
```

####SIMULATING A COALESCENT TREE 
```{r}
coal.tree<-rcoal(n=100)
plotTree(coal.tree,ftype="off")
coal.obj<-ltt(coal.tree,plot=FALSE)
print(coal.obj)
```

####PRINT COAL LTT PLOT
```{r}
obj<-ltt(coal.tree,log.lineages=FALSE,log="y")
```

####COMPARING GAMMA BETWEEN PURE BIRTH AND COALESCENT SIMULATED TREES 
```{r}
darter.gamma <- obj$gamma
darter.gamma
trees<-pbtree(n=100,nsim=200,scale=max(nodeHeights(coal.tree)))
ltt95(trees,log=TRUE)
title(main="Simulated coalescent trees compared to pure-birth LTTs")
ltt(coal.tree,add=TRUE,log.lineages=FALSE,col="red",lwd=2,lty="dashed")
```
The coalescent tree shows an exponential growth even when logged

##Incomplete Sampling
Authors were missing 15 species
Going to see what affect that has by generating 216 species and subsampling down to 201 and examining each gamma. If ours is extreme in comparison, we can conclude it's not caused by incomplete sampling

####SIMULATE PURE BIRTH TREE
```{r}
library(geiger)
age <- 25.91862
richness <- 216
darterbirth =  (log(richness) - log(2))/age ; darterbirth
```

####SIMULATE INCOMPLETE SAMPLING
```{r}
richness <- 216
missing <- 15
num_simulations<-200 
g1_null<-numeric(num_simulations) #g1_null will hold the simulated gamma values
for(i in 1:num_simulations) {
  sim.bdtree(darterbirth, d=0, stop = "taxa", n=richness)->sim_tree 
  drop.random(sim_tree, missing)->prune # prune down to the # of taxa in the phylogeny
  gammaStat(prune)->g1_null[i]
}
```

####PLOT HISTOGRAM OF NULL DISTRIBUTION AND ADD ARROW TO SHOW OBSERVED GAMMA
```{r}
hist(g1_null)
arrows(darter.gamma, 40, darter.gamma, 0, col="red", lwd=2) 
```

####COMPARE OBSERVED GAMMA TO SIMULATED VALUES
```{r}
smallerNull<-g1_null<=darter.gamma
count<-sum(smallerNull)
```

####USE TO GENERATE P VALUE 
```{r}
mccr_pval<-(count+1)/(num_simulations+1)
mccr_pval
```

The gamma value is definitely just due to subsampling

