---
title: "Phylogenetically Independent Constrasts"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##PACKAGES
```{r}
library(ape)
library(phytools)
```


##CHALLENGE PROBLEM 1
Fit a OLS (non-phylogenetic) regression model for y~x and a phylogenetic regression model using contrasts. What do you find? Why?  
  
OLS regression model:
```{r}
obj<-read.csv("pic-exercise-data.csv",row.names=1); head(obj)
fit.ols<-lm(y~x,data=obj); fit.ols
summary(fit.ols)
plot(obj[,c("x","y")])
abline(fit.ols,lwd=2,lty="dashed",col="red")
```

Phylogenetic regression model using contrasts:
```{r}
ex.tree<-read.tree("pic-exercise-tree.tre")
x.data<-setNames(obj[,"x"],rownames(obj))
y.data<-setNames(obj[,"y"],rownames(obj))
pic.x<-pic(x.data,ex.tree) #categorize buccal length by phylogeny
pic.y<-pic(y.data,ex.tree) #categorize gape width by phylogeny
fit.pic<-lm(pic.y~pic.x+0) #fit linear model with phylogeny considered
fit.pic
```
```{r}
summary(fit.pic)
```

Plot linear model over scatterplot with phylogeny considered:
```{r}
plot(pic.x,pic.y,xlab="PICs for x",ylab="PICs for y")
abline(fit.pic,lwd=2,lty="dashed",col="red")
```

We find that, without taking into account phylogentic history, x seems to have no effect on y. Once the tree is used to weight the points by relatedness, a linear relationship between x and y does emerge. Not taking into account phylogeny apparently can lead to making a type II error as well as the possibility in other situations of causing a type I error.

##TUTORIAL
####Linear Model Basics
Fitting linear regressions to first gape width and buccal length. 
View Data:
```{r}
obj<-read.csv("Centrarchidae.csv",row.names=1); head(obj)
```

Plot effect of buccal length on gape width
```{r}
plot(obj[,c("buccal.length","gape.width")])
```

Fit OLS regression model:
```{r}
fit.ols<-lm(gape.width~buccal.length,data=obj); fit.ols
```
```{r}
summary(fit.ols)
```

Add OLS regression line to graph:
```{r}
plot(obj[,c("buccal.length","gape.width")])
abline(fit.ols,lwd=2,lty="dashed",col="red")
```

Fit PIC to take into account phylogeny:
```{r}
cent.tree<-read.tree("Centrarchidae.tre")
buccal.length<-setNames(obj[,"buccal.length"],rownames(obj))
gape.width<-setNames(obj[,"gape.width"],rownames(obj))
pic.bl<-pic(buccal.length,cent.tree) #categorize buccal length by phylogeny
pic.gw<-pic(gape.width,cent.tree) #categorize gape width by phylogeny
fit.pic<-lm(pic.gw~pic.bl+0) #fit linear model with phylogeny considered
fit.pic
```
```{r}
summary(fit.pic)
```

Plot linear model over scatterplot with phylogeny considered
```{r}
plot(pic.bl,pic.gw,xlab="PICs for buccal length",ylab="PICs for gape width")
abline(fit.pic,lwd=2,lty="dashed",col="red")
```

####Simulating a Tree to Test PIC
Simulate a coalescent tree:
```{r}
set.seed(21)
## simulate a coalescent shaped tree
tree<-rcoal(n=100)
plotTree(tree,ftype="off")
```

Simulate uncorrelated Brownian evolution:
```{r}
x<-fastBM(tree) #simulate trait 1
y<-fastBM(tree) #simulate trait 2
par(mar=c(5.1,4.1,2.1,2.1)) #set plot dimensions
plot(x,y)
fit<-lm(y~x) #fit linear regression to data
abline(fit,lwd=2,lty="dashed",col="red") #add lm line
```

Return descriptions of fit:
```{r}
fit
```
```{r}
summary(fit)
```
```{r}
anova(fit)
```

Observe that phylogeny can very easily cause a type I error. The close taxonomical relationship of the data points within the two clusters causes the traits to appear correlated even when they are not.  
Visualize the relatedness of the data points in the scatterplot:
```{r}
phylomorphospace(tree,cbind(x,y),label="off",node.size=c(0.5,0.7))
abline(fit,lwd=2,lty="dashed",col="red")
```

Use Felsenstein's algorithim to weight the contrasts between species:
```{r}
ix<-pic(x,tree)
iy<-pic(y,tree)
fit<-lm(iy~ix-1) ## we have to fit the model without an intercept term
fit
```
```{r}
summary(fit)
```
```{r}
anova(fit)
```