---
title: "Drift Exercise"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Question 1
The expected number of successes based on the formula from class would the be same as the initial propotion of successes. In other words, the proportion of successes is no *expected* to change between generations so the expected number of successes would be _10 * 0.01_ which is equal to 1.  
Simulation:
```{r}
n<-100
N<-50
p<-0.1
sum(rbinom(n,2*N,p))/(2*N)/n
```
The simulation 

##Question 2
```{r}
n<-1 ; N<-50 ; p<-0.1
T<-5
L<-3

driftFun<-function(T,L){
  freqs<-matrix(nrow = T, ncol = L)
  freqs[1,]<-rep(p,L)
  for(i in 2:T){
    for(j in 1:L){
      freqs[i,j]<-rbinom(1,2*N,freqs[i-1,j])/(2*N)
    }
  }
  return(freqs)
}

freqsNew<-driftFun(5,3) ; freqsNew
```


##Question 3
```{r}
n<-1 ; N<-100 ; p<-0.1
T<-1000
L<-10000
freqs3<-driftFun(T,L)

#(a)
numDis<-sum(freqs3[T,] == 0) ; numDis
probDis<-numDis/(L) ; probDis
#(b)
numFix<-sum(freqs3[T,] == 1) ; numFix
probFix<-numFix/(L) ; probFix
#(c)
numNot<-(L)-numDis-numFix ; numNot
probNot<-numNot/(L) ; probNot
#(d)
plot(1, type="n", xlab="Generation", ylab="Frequency", main = "N=100", xlim=c(0, T), ylim=c(0, 1))
for (i in 1:100){
  lines(freqs3[,i],type = "l",col=i)
}
```

New simulation (p=0.6):
```{r}
n<-1 ; N<-100 ; p<-0.6
T<-1000
L<-10000
freqs3b<-driftFun(T,L)
#(e)
numDis<-sum(freqs3b[T,] == 0) ; numDis
probDis<-numDis/(L) ; probDis
#(f)
numFix<-sum(freqs3b[T,] == 1) ; numFix
probFix<-numFix/(L) ; probFix
#(g)
numNot<-(L)-numDis-numFix ; numNot
probNot<-numNot/(L) ; probNot
```

##Question 4
```{r}
n<-1 ; p<-0.1 ; T<-1000 ; L<-10000
N<-10 ; freqs10<-driftFun(T,L)
N<-500 ; freqs500<-driftFun(T,L)
N<-1000 ; freqs1000<-driftFun(T,L)
```


##Question 5
```{r}
#(a)
par(mfrow=c(2,2))
plot(1, type="n", xlab="Generation", ylab="Frequency", main = "N=10", xlim=c(0, T), ylim=c(0, 1))
for (i in 1:100){
  lines(freqs10[,i],type = "l",col=i)
}
plot(1, type="n", xlab="Generation", ylab="Frequency", main = "N=500", xlim=c(0, T), ylim=c(0, 1))
for (i in 1:100){
  lines(freqs500[,i],type = "l",col=i)
}
plot(1, type="n", xlab="Generation", ylab="Frequency", main = "N=1000", xlim=c(0, T), ylim=c(0, 1))
for (i in 1:100){
  lines(freqs1000[,i],type = "l",col=i)
}
#(c)
numFix10<-sum(freqs10[T,] == 1) ; numFix10
probFix10<-numFix10/(L) ; probFix10
numFix500<-sum(freqs500[T,] == 1) ; numFix500
probFix500<-numFix500/(L) ; probFix500
numFix1000<-sum(freqs1000[T,] == 1) ; numFix1000
probFix1000<-numFix1000/(L) ; probFix1000
```


##Question 6


##Question 7


##Question 8


##Question 9


##Question 10


##Question 11
